The Magnetorotational Instability Prefers Three Dimensions

### 2019-11-11 paper/paper.tex
fixed sign error in one equation. results are still just fine.
i keep finding little things to improve. i'll stop soon.
ok. last tiny edit
New section on analysis on 3D -> 2D transition.

@@

\Phi(x) \ = \ \frac{\pi^{2}}{d^{2}} + \eps^{2} \frac{B_{0}^2 f^2 \Ro\, \left(k_{z}^2 R+\Ro\, k_{y} ^2\right) - B_{0}^4 k^4 +f^2 (\Ro+1) (\omega -f\, \Ro\, x\, k_{y}
   )^2  }{B_{0}^4 k_{z}^2}.

@@

\psi_{2}''(x) + \frac{\pi^{2}}{d^{2}}  \psi_{2}(x) + \Phi_{2}(x)\psi_{0}(x)\ = \ 0.

@@

\sigma^{2}\ = \ - \omega^{2} \ = \  \frac{\Ro\,B_{0}^{2}}{(\Ro+1)}\left[ k_{z}^{2}\left(R-\frac{d^2}{\pi^{2}}k_{z}^{2}\right)+\Upsilon\,k_{y}^{2}\right]+\ldots,

@@

\Upsilon \ = \  \frac{\left(\pi^{2}+6\right) \Ro+\pi^{2}-6}{12} \ \approx \ 1.31\quad\text{for}\quad{\Ro} \ = \ 0.75,

@@

>>>>

\newcommand{\s}{\text{s}}
\newcommand{\x}{\text{x}}
\newcommand{\y}{\text{y}}

@@

\Phi(x) \ = \ -\frac{\pi^{2}}{d^{2}} - \frac{\pi^{2} \eps^{2}}{d^{2}}\left[R - \frac{d^2 }{\pi^2}k_z^2 + \frac{\Ro^2 B_0^2  k_y^2+(1+\Ro) (\omega + f \Ro\, x k_y) ^2}{ \Ro\, B_0^2
   k_z^2} \right].

@@

\psi_{2}''(x) + \frac{\pi^{2}}{d^{2}}  \psi_{2}(x) \ = \  \Phi_{2}(x)\psi_{0}(x) .

@@

\sigma^{2}\ = \ - \omega^{2} \ = \  \frac{\Ro\,B_{0}^{2}}{1+\Ro}\left[ k_{z}^{2}\left(R-\frac{d^2}{\pi^{2}}k_{z}^{2}\right)+\Upsilon\,k_{y}^{2}\right]+\ldots,

@@

\Upsilon \ = \  \Ro + (1+ \Ro) \frac{\pi^{2}-6}{12}  \ \approx \ 1.31\quad\text{for}\quad{\Ro} \ = \ 0.75,

@@

\subsection{Analysis of the transition to 2D}
\label{sec:2D_transition}

We can understand the transition from three- to two-dimensional instability semi-analytically. We do this by conducting a similar analysis as the above asymptotic calculation, only close to the fastest growing 2D modes. We can look for values for $\SSC$ where the analogous value of $\Upsilon(\SSC) = 0$.

The calculation becomes very messy. We therefore only sketch the details, and report the main results.
We analyse the growth rate as a function of  $k_{y},k_{z}$ and $\SSC$ near $k_{y} \approx0$.  We therefore expand the whole problem in a power-series ansatz
\Beq
\gamma(\SSC,k_{y},k_{z}) \ = \   \sum_{n\ge 0} \gamma_{n}(\SSC,k_{z}) k_{y}^{n}. 
\Eeq
We use the same asymptotic techniques as before to obtain each progressive correction. We solve the leading order by virtue that all non-constant coefficients vanish for 2D modes. The following dispersion relations determine the maximum growth rate and the corresponding wavenumber,
\Beq
\mathfrak{D}_{0}(k_{z}) &=& \frac{f k_z^2 \left(B_0^2 S k_z^2+\gamma _0^2 (f+S)\right)}{\left(B_0^2 k_z^2+\gamma _0^2\right){}^2}+\frac{\pi ^2}{d^2}+k_z^2 \ = \ 0 \\
\frac{\mathfrak{D}_{0}'(k_{z})}{2k_{z}} &=& \frac{\gamma _0^2 f \left(B_0^2 k^2 (S-f)+\gamma _0^2 (f+S)\right)}{\left(B_0^2 k^2+\gamma _0^2\right)^3}+1 \ = \ 0.
\Eeq
We can transform the these relations into a polynomials in terms of non-dimensional variables,
\Beq
\label{dispersion}
 \Ro^2 \s^2 \y^2+   (2 \Ro + \s - \Ro \, \s + \Ro^2 \s \y)\s  \y \x + (1 - \s + 2 \Ro\, \s \, \y) \x^2 +\x^3 \ = \ 0 \\ 
(2 \Ro + \s - \Ro \, \s + \Ro^2 \s \y) \s  \y + 2(1 - \s + 2 \Ro\, \s \, \y) \x +3\x^2 \ = \ 0
\label{critical-dispersion}
\Eeq
where
\Beq
\x  \ = \ \frac{d^2 k_z^2}{\pi ^2}, \quad  \quad \y \ = \ \frac{\gamma_{0}^{2}}{S^{2}}, \quad \quad \s  \ = \ \frac{S}{S_{c}}.
\Eeq
These are directly related to the parameters reported on figure~\ref{fig:growth_rate}.
The two polynomials are straightforward to solve numerically for $\x,\y$ given $\s, \Ro$. However we  still want to find critical values for $\s$, where the fastest growing modes are two dimensional. From now on, the calculation becomes sufficiently complicated that we only outline the methodology.

We find the higher-order $\gamma_{n}(\s,k_{z})$ coefficients by applying order-by-order solvability conditions. We first find that $\gamma_{1}(\s, k_{z}) = 0$ identically. This renders the first-order correction problem solvable, which means solving it and moving on the the 2nd-order corrections. The structure of the growth rate is therefore 
\Beq
\gamma(\s,k_{y},k_{z}) \ = \ \gamma_{0}(\s,\x) + \gamma_{2}(\s,\x) k_{y}^{2} \ + \ \ldots 
\Eeq
This is the same structure as the asymptotic calculation in the previous section. Now effectively $\Upsilon \propto \gamma_{2}$. The case with dominant 2D modes corresponds to $\gamma_{2}(\s,\x)=0$. Finding this condition is a formidable calculation, which produces the polynomial expression  
\Beq
\Ro^6 \s^6 \y^6 + \Ro^4 \s^5 \y^4 (3 \xi  (\Ro-1) \s+\Ro (6 \y-1)) \x + \nonumber \\  
\frac{\Ro^2 \s^4 \y^3}{12}  \left(12 \Ro^2 (15 \y-4)-3 (4 \xi -3)
   (\Ro-1)^2 \s^2+24 \xi  (4 \Ro+1) \Ro \, \s\right)\x^{2} + \nonumber \\ 
\frac{\Ro\, \s^3 \y^2}{6}  \left(-3 (4 \xi -3) \left(\Ro^2-1\right) \s^2+12
   \Ro^2 (10 \y-3)+36 \xi  (\Ro+2) \Ro\, \s\right)\x^{3} +\nonumber \\ 
   \left(\Ro^2 \s^2 \y (15 \y-4)-\frac{1}{4} (4 \xi -3) (\Ro+1)^2 \s^4 \y+6 \xi 
   \Ro\, \s^3 \y\right) \x^{4} +\nonumber \\ 
   -\s (\xi  \Ro \, \s-6 \Ro \, \y+\Ro+\xi  \s) \x^{5} +  
   \x^{6} \ = \ 0.
   \label{2D-condition}
\Eeq
The new coefficient is
\Beq
\xi \ = \ \frac{\pi^{2}-6}{12} \ = \ \frac{\int_{-d/2}^{d/2} x^2 \cos ^2\left(\frac{\pi  x}{d}\right) \dd{x}}{\frac{d^{2}}{\pi^{2}}\int_{-d/2}^{d/2}  \cos ^2\left(\frac{\pi  x}{d}\right) \dd{x}}.
\Eeq
This results from projecting the squared shear onto the leading-order eigenmodes, just as in the previous section.

Given $\Ro$, we solve \eq{dispersion}, \eq{critical-dispersion}, and  \eq{2D-condition} for $\x,\y,\s$ using Newton's method.

\begin{center}
\begin{tabular}{cccccccccccccc}
\textbf{$\Ro$} & \textbf{$k_{z} d/\pi$}  & \textbf{$\gamma_{\max}/|S_{2D}|$}  &  \textbf{$S_{2D}/S_{c}$} \\
                   1.00  & 0.708276 & 0.209120 &   2.13892 \\ 
                   0.75 & 0.682819 & 0.214895 & 2.05136 \\
                   0.50 & 0.651764 & 0.220102 & 1.95000 \\
                   0.25 & 0.615396 & 0.224939 & 1.83881 \\
                   0.10 & 0.593345 & 0.229047 & 1.77526
\end{tabular}
\end{center}
The numbers for $\Ro = 0.75$ match the numerical results shown in Figure~\ref{fig:growth_rate} excellently.

Finally, we can continue with the calculation conceptually. Because of symmetry $\gamma_{3}(\s,\x) = 0$ identically. This means that to leading order 
\Beq
\gamma(\s,k_{y},k_{z}) \ = \ \gamma_{0}(\s,\x) + \gamma_{2}(\s,\x) k_{y}^{2}  + \gamma_{4}(\s,\x) k_{y}^{4} \ + \ \ldots
\Eeq 
For $\SSC > S_{2D}/S_{c}$, $\gamma_{2} > 0$. This is sufficient to determine the most unstable 3D wavenumber near the transition to dominant 2D modes, 
\Beq
k_{y}^{2} \ \approx \ -\frac{\gamma _2}{2 \gamma_4} \ \propto \  S_{2D}/S_{c} - \SSC . 
\Eeq 
Generically $\gamma_{4} \ne 0$. We assume that $\gamma_{4} < 0$, but we do not know for sure if this is the case. The calculation is considerably more challenging that what we already have done. If there is a negative 4th-order feedback, then 
\Beq
k_{y} \ \sim \ \sqrt{S_{2D}/S_{c} - \SSC}. 
\Eeq
This is consistent with Figure~\ref{fig:growth_rate}d near the critical point. If $\gamma_{4} > 0$, then the calculation would proceeded to 6th order (and so on), and the critical exponent will be smaller. Nevertheless the general picture will stay the same.
